---
title: "Chapter 12"
format: 
    html:
    theme: cosmo
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(brms)
library(ggthemes)

theme_set(
  theme_hc() +
  theme(axis.ticks.y = element_blank(),
        plot.background = element_rect(fill = "grey92"))
)
```


Zeit paar Monster zu erzeugen, indem man Modelle mixt

- over-dispersion
 Erweiterung von binomial/poisson um sources of unkown variance zu lindern
 
 - zero-inflated/augmented
 Binary event + GLM likelihood
 
 - ordered categorical model
 für Modelle mit einem fix geordneten Outcome und einem spezillere Link-function:
 dem kumulativen link
 
 
## over dispersion counts
 If counts come from more than one distribution it leads to higher amounts of variation than the model expects. The model becomes to excited
 
This is known as over-dispersion (dispersion = other word for variance)

continouse mixture models: linear model is attached not to the observations themselves but rather to a distribution of observations

### beta binomial model
Assumes each count to have it's own probability of success

$$
\operatorname{Beta}(y | \alpha, \beta) = \frac{y^{\alpha - 1} 
(1 - y)^{\beta - 1}}{\text B (\alpha, \beta)}
$$


```{r}
pbar  <- .5
theta <- 5

ggplot(data = tibble(x = seq(from = 0, to = 1, by = .01)),
       aes(x = x, y = rethinking::dbeta2(x, prob = pbar, theta = theta))) +
  geom_area(fill = "plum4") +
  scale_x_continuous("probability space", breaks = c(0, .5, 1)) +
  scale_y_continuous(NULL, breaks = NULL) 
```

$$
\operatorname{A_i} \sim \operatorname{BetaBinomial}(N,\bar{p},\theta) \\
logit(\bar{p_i}) = \alpha_{GID[i]} \\
\alpha_j \sim Normal(0,1.5) \\
\theta = \phi + 2 \\
\phi \sim Exponential(1)
$$

```{r}
data(UCBadmit, package = "rethinking") 
d <- 
  UCBadmit |>
  mutate(gid = ifelse(applicant.gender == "male", "1", "2"))
rm(UCBadmit)

beta_binomial2 <- custom_family(
  "beta_binomial2", dpars = c("mu", "phi"),
  links = c("logit", "log"),
  lb = c(0, 2), ub = c(1, NA),
  type = "int", vars = "vint1[n]"
)

stan_funs <- "
  real beta_binomial2_lpmf(int y, real mu, real phi, int T) {
    return beta_binomial_lpmf(y | T, mu * phi, (1 - mu) * phi);
  }
  int beta_binomial2_rng(real mu, real phi, int T) {
    return beta_binomial_rng(T, mu * phi, (1 - mu) * phi);
  }
"

stanvars <- stanvar(scode = stan_funs, block = "functions")

b12.1 <-
  brm(data = d, 
      family = beta_binomial2,  
      admit | vint(applications) ~ 0 + gid,
      prior = c(prior(normal(0, 1.5), class = b),
                prior(exponential(1), class = phi)),
      iter = 2000, warmup = 1000, cores = 4, chains = 4,
      stanvars = stanvars,  
      seed = 12,
      file = "fits/b12.01")

print(b12.1)

post <- as_draws_df(b12.1)

library(tidybayes)
post |> 
  transmute(da = 
              b_gid1 - b_gid2) |> 
  mean_qi(.width = .89) |> 
  mutate_if(is.double, round, digits = 3)

set.seed(12)

lines <-
  post |>
  mutate(p_bar = inv_logit_scaled(b_gid2)) |>
  slice_sample(n = 100) |>
  select(.draw, p_bar, phi) |>
  expand_grid(x = seq(from = 0, to = 1, by = .005)) |>
  mutate(density = pmap_dbl(list(x, p_bar, phi), rethinking::dbeta2))

str(lines)

lines |> 
  ggplot(aes(x = x, y = density)) + 
  stat_function(fun = rethinking::dbeta2,
                args = list(prob = mean(inv_logit_scaled(post %>% pull(b_gid2))),
                            theta = mean(post %>% pull(phi))),
                linewidth = 1.5, color = canva_pal("Green fields")(4)[4]) +
  geom_line(aes(group = .draw),
            alpha = .2, color = canva_pal("Green fields")(4)[4]) +
  scale_y_continuous(NULL, breaks = NULL, limits = c(0, 3)) +
  labs(subtitle = "distribution of female admission rates",
       x = "probability admit")

data(Kline, package = "rethinking")
d <- 
  Kline %>% 
  mutate(p          = rethinking::standardize(log(population)),
         contact_id = ifelse(contact == "high", 2L, 1L),
         cid        = contact)
rm(Kline)

print(d)
```

$$
\begin{align*}
\text{total_tools}_i & \sim \operatorname{Gamma-Poisson} (\mu, \alpha) \\
\text{log}(\mu) & = \beta_0 \\
\beta_0         & \sim \operatorname{Normal}(3, 0.5) \\
\alpha          & \sim \operatorname{Gamma}(0.01, 0.01)
\end{align*}
$$

```{r}
ggplot(data = tibble(x = seq(from = 0, to = 60, by = .1)),
       aes(x = x, y = dgamma(x, shape = 0.01, rate = 0.01))) +
  geom_area(color = "transparent", 
            fill = "pink") +
  scale_x_continuous(NULL) +
  scale_y_continuous(NULL, breaks = NULL) +
  coord_cartesian(xlim = c(0, 50)) +
  ggtitle(expression(brms~default~gamma(0.01*", "*0.01)~shape~prior))

b12.2a <-
  brm(data = d, 
      family = negbinomial,
      total_tools ~ 1,
      prior = c(prior(normal(3, 0.5), class = Intercept),  # beta_0
                prior(gamma(0.01, 0.01), class = shape)),  # alpha
      iter = 2000, warmup = 1000, cores = 4, chains = 4,
      seed = 12,
      file = "fits/b12.02a")

print(b12.2a)
```

## zero inflated outcomes

Very often, the things we can measure are not emissions from any pure process. Instead, they are mixtures of multiple processes. Whenever there are different causes for the same observation, then a mixture model may be useful. A mixture model uses more than one simple probability distribution to model a mixture of causes. In effect, these models use more than one likelihood for the same outcome variable

```{r}
prob_drunk <- .2
rate_work <- 1

n <- 365

set.seed(99)
drunk <- rbinom(n,
                size = 1,
                prob = prob_drunk)

y <- (1-drunk) * rpois(n,rate_work)

tibble(
  drink = factor(drunk, levels = 1:0),
  y     = y
) -> d

d |> 
  ggplot(aes(x = y)) +
  geom_histogram(aes(fill = drink),
                 binwidth = 1, linewidth = 1/10, color = "grey92") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  xlab("Manuscripts completed") +
  theme(legend.position = "none")

b12.3 <- 
  brm(data = d, 
      family = zero_inflated_poisson,
      y ~ 1,
      prior = c(prior(normal(1, 0.5), class = Intercept),
                prior(beta(2, 6), class = zi)),  
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 12,
      file = "fits/b12.03") 

print(b12.3)
```

## Orderedcategoricaloutcomes

```{r}
data(Trolley, package = "rethinking")
d <- Trolley
rm(Trolley)

p1 <-
  d |> 
  ggplot(aes(x = response, fill = after_stat(x))) +
  geom_histogram(binwidth = 1/4, linewidth = 0) +
  scale_fill_gradient(low = canva_pal("Green fields")(4)[4],
                      high = canva_pal("Green fields")(4)[1]) +
  scale_x_continuous(breaks = 1:7) +
  theme(axis.ticks = element_blank(),
        axis.title.y = element_text(angle = 90),
        legend.position = "none")
p1

p2 <-
  d %>%
  count(response) %>%
  mutate(pr_k     = n / nrow(d),
         cum_pr_k = cumsum(pr_k)) %>% 
  
  ggplot(aes(x = response, y = cum_pr_k, 
             fill = response)) +
  geom_line(color = canva_pal("Green fields")(4)[2]) +
  geom_point(shape = 21, color = "grey92", 
             size = 2.5, stroke = 1) +
  scale_fill_gradient(low = canva_pal("Green fields")(4)[4],
                      high = canva_pal("Green fields")(4)[1]) +
  scale_x_continuous(breaks = 1:7) +
  scale_y_continuous("cumulative proportion", 
                     breaks = c(0, .5, 1), limits = c(0, 1)) +
  theme(axis.ticks = element_blank(),
        axis.title.y = element_text(angle = 90),
        legend.position = "none")
p2
```


```{r}
library(rethinking)
data("Trolley")
d <- Trolley

dim(d)

simplehist(d$response, xlim = c(1,7), xlab = "Response")

d |> 
  ggplot(aes(response)) +
  geom_bar() +
  scale_x_continuous(breaks = c(1:7))

pr_k <- table(d$response) / nrow(d)

cum_pr_k <- cumsum(pr_k)

plot(1:7, cum_pr_k, type = "b", xlab = "response",
     ylab = "cumulative proportion", ylim = c(0,1))
```

$$
log \frac{Pr(y_i\leq k)}{1-Pr(y_i\leqk)} = \alpha_k
$$

```{r}
logit <- function(x) {
  log(x/(1-x))
}

round(lco <- logit(cum_pr_k),2)

plot(1:7, lco, type = "b", xlab = "response",
     ylab = "cumulative proportion")
```


Formula for the likelihood:

$$
p_k = Pr(y_i=k) = Pr(y_i \leq k) - Pr(y_i \leq k-1)
$$

Daraus ergibt sich das Modell

$$
R_i \sim Ordered-logit(\phi_i,\kappa) \\
\phi_i = 0 \\
\kappa_k \sim Normal(0,1.5)
$$

```{r}
dat <- list(
R = d$response,
    A = d$action,
    I = d$intention,
    C = d$contact )

m12.5 <- ulam( alist(
R ~ dordlogit( phi , cutpoints ),
phi <- bA*A + bC*C + BI*I ,
BI <- bI + bIA*A + bIC*C , 
c(bA,bI,bC,bIA,bIC) ~ dnorm( 0 , 0.5 ),
cutpoints ~ dnorm( 0 , 1.5 )) , data=dat , chains=4 , cores=4 ) 
precis( m12.5 )
```

```{r}
plot(NULL, type = "n", xlab = "intention", ylab = "probability",
     xlim = c(0,1), ylim = c(0,1), xaxp = c(0,1,1), yaxp=c(0,1,2))

kA <- 0
kC <- 0
kI <- 0:1
pdat <- data.frame(A=kA,C=kC,I = kI)
phi <- link(m12.5, data = pdat)$phi

post <- extract.samples(m12.5)
for (s in 50) {
  pk <- pordlogit(1:6, phi[s,], post$cutpoints[s,])
  for (i in 1:6) lines(kI, pk[,i], col = "red4")
}

```

## ordered categorical predictors

```{r}
levels(d$edu)
edu_levels <- c(6,1,8,4,7,2,5,3)
d$edu_new <- edu_levels[d$edu]

library(gtools)
set.seed(1805)
delta <- rdirichlet(10, alpha = rep(2, 7)) 

str(delta)
delta %>% 
  data.frame() %>%
  set_names(1:7) %>% 
  mutate(row = 1:n()) %>% 
  pivot_longer(-row, names_to = "index") %>% 
  
  ggplot(aes(x = index, y = value, group = row,
             alpha = row == 3, color = row == 3)) +
  geom_line() +
  geom_point() +
  scale_alpha_manual(values = c(1/3, 1)) +
  scale_color_manual(values = canva_pal("Green fields")(4)[1:2]) +
  ylab("probability") +
  theme(legend.position = "none")
```

```{r}
b12.6 <- 
  brm(data = d, 
      family = cumulative,
      response ~ 1 + action + contact + intention + mo(edu_new),  # note the `mo()` syntax
      prior = c(prior(normal(0, 1.5), class = Intercept),
                prior(normal(0, 1), class = b),
                # note the new kinds of prior statements
                prior(normal(0, 0.143), class = b, coef = moedu_new),
                prior(dirichlet(2, 2, 2, 2, 2, 2, 2), class = simo, coef = moedu_new1)),
      iter = 2000, warmup = 1000, cores = 4, chains = 4,
      seed = 12,
      file = "fits/b12.06")
```

